A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-02-20 paper/bare_conf.tex
Improve code transformation description. Add splitting and snippets

Local assembly consists of evaluating so called element stiffness matrix and vector; in this work, we focus on computation of matrices, which is the costly part of the process. A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain. Quadrature algorithms (based on numerical integration), which follow on from the mathematical description of a stiffness matrix, are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature 2}. The FFC compiler used by Firedrake and Fenics can generate assembly kernels using optimized quadrature schemes. It is worth noticing, however, that problems and considerations similar to that exposed below can be found also in non-FFC assembly codes.

@@

@@

% Register tiling in its two forms
One notable problem assembly kernels are exposed to concerns register allocation and register locality. The critical situation occurs when loop trip counts and number of accessed matrices along the $j$ and $k$ dimensions are such that available vector registers are under-utilized. Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management in order to maximize performance. Canonical optimizations (i.e. non domain-driven), such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem. In our compiler we support these optimizations either by means of explicit source code transformation (interchange, register tiling, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll). With register tiling, in particular, we slice the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and we apply unrolling to individual blocks.

@@

@@

% Splitting

>>>>

Local assembly consists of evaluating so called element stiffness matrix and vector; in this work, we focus on computation of matrices, which is the costly part of the process. A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain. Quadrature algorithms (based on numerical integration), which follow on from the mathematical description of a stiffness matrix, are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature2}. The FFC compiler used by Firedrake and Fenics can generate assembly kernels using optimized quadrature schemes. It is worth noticing, however, that problems and considerations similar to that exposed below can be found also in non-FFC assembly codes.

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment and $licm$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$; in general, this could not be the case.}
\label{code:helmholtz-licm}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment, $licm$ and $split$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, the split factor is 2.}
\label{code:helmholtz-split}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

% Register allocation and register locality
One notable problem assembly kernels are exposed to concerns register allocation and register locality. The critical situation occurs when loop trip counts and number of accessed matrices along the inner dimensions are such that available vector registers are under-utilized. Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management in order to maximize performance. Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem. In our compiler we support these optimizations either by means of explicit source code transformation (interchange, register tiling, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll). With register tiling, in particular, we slice the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and we apply unrolling to individual blocks to expose ILP.

@@

% Splitting
In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging. If the number of variables independent of the innermost-loop dimension, i.e. basis function matrices, temporaries introduced by $licm$ and constants, is larger than the available registers on the CPU, it is likely to obtain poor register reuse. For example, applying $licm$ to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers. One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the simpler Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}. Splitting an expression has, in general, several drawbacks. Firstly, it increases the number of accesses to $A$ proportially to the ``split factor'', which is the number of sub-expressions produced. Secondly, depending on how the split is executed, it can lead to redundant computation (e.g. the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}). Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same matrix could be accessed in different sub-expressions, requiring a proportional number of loads be performed. Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself are based upon heuristics that aim at minimizing the aforementioned issues.

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=2$ elements). The padding, data alignment, $licm$ and $op$-$vect$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, the unroll-and-jam factor is 1.}
\label{code:helmholtz-split}
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// Call $Load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}
